---
title: "Final Project"
author: "Martina Radoslavov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(GGally)
library(gtsummary)
```

## Load and Clean Data
```{r Load Data}
GliomaData <- read.csv('GliomaData/TCGA_GBM_LGG_Mutations_all.csv')
head(GliomaData)

GliomaINFO <- read.csv('GliomaData/TCGA_InfoWithGrade.csv')
head(GliomaINFO)
```
```{r}
unique(GliomaINFO$Race)
```

```{r Exploration}
unique(GliomaData[, c("Grade", "Primary_Diagnosis")])
```

```{r DELETE}
any(is.na(GliomaData))
GliomaData[GliomaData == "MUTATED"] <- 1
GliomaData[GliomaData == "NOT_MUTATED"] <- 0
head(GliomaData)
```
There are no missing data. 

## Distribution/histogram Plots 
### normal distribution? 

## Box plots 


```{r Heatmap}
corr_mat <- cor(GliomaINFO, use = "pairwise.complete.obs")

ggcorr(GliomaINFO,
       label = TRUE,
        label_size = 2,
        hjust = 1,
        size = 2,
        layout.exp = 2)
```

```{r}
tbl_summary(GliomaData %>% select(-c(Case_ID, Age_at_diagnosis, Project, Primary_Diagnosis)), by = Grade)
# Add p-value to tbl
```

## ESTABLISH COLINEARITY
```{r}
threshold <- 0.5

high_corr_pairs <- which(abs(corr_mat) > threshold & lower.tri(corr_mat), arr.ind = TRUE)

pairs_df <- data.frame(
  Var1 = rownames(corr_mat)[high_corr_pairs[,1]],
  Var2 = colnames(corr_mat)[high_corr_pairs[,2]],
  Correlation = corr_mat[high_corr_pairs]
)

pairs_df
```
Because ATRX has a stronger correlation with Grade then TP53 would, we will remove TP53 from the model. 

```{r}
muts <- GliomaData %>% 
  select(IDH1:PDGFRA) %>%
  colnames()

for (m in muts) {
  cat("\n===== Testing:", m, "=====\n")
  
  tab <- table(GliomaData$Grade, GliomaData[[m]])
  print(tab)
  
  exp_counts <- chisq.test(tab)$expected
  cat("\nExpected counts:\n")
  print(round(exp_counts, 2))
  
  # Check assumption
  if (any(exp_counts < 1) || mean(exp_counts < 5) > 0.2) {
    cat("\n Chi-square assumptions NOT met → using Fisher's exact test.\n")
    print(fisher.test(tab))
  } else {
    cat("\n️ Chi-square assumptions OK.\n")
    print(chisq.test(tab))
  }
}
```
The following mutations do not show a significant difference between the two Glioma grades: PIK3CA, BCOR, CSMD3, FAT4
We can omit them from the clustering model. 

## Logistic Regression Model 
```{r}
removee <- c('PIK3CA', 'BCOR', 'CSMD3', 'FAT4', 'TP53')

mData <- GliomaINFO %>%
  select(-all_of(removee)) 
```

```{r}
model <- glm(Grade ~ .,
             data = mData,
             family = binomial)
summary(model)

```


