---
title: "Final Project"
author: "Martina Radoslavov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(GGally)
library(gtsummary)
```

## Load and Clean Data
```{r Load Data, results=FALSE}
GliomaData <- read.csv('GliomaData/TCGA_GBM_LGG_Mutations_all.csv')
head(GliomaData)

GliomaINFO <- read.csv('GliomaData/TCGA_InfoWithGrade.csv')
head(GliomaINFO)
```


```{r Exploration}
unique(GliomaData[, c("Grade", "Primary_Diagnosis")])
```
There are no missing data. 

```{r Number of LGG vs GBM Cases}
GliomaINFO %>% 
  count(Grade) %>% 
  mutate(Grade = factor(Grade, levels = c(0, 1),
                        labels = c("LGG", "GBM"))) %>%
  ggplot(aes(x = Grade, y = n, fill = Grade)) +
  geom_col(width = 0.55, color = "white") +
  geom_text(aes(label = n),
            vjust = -0.3, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("#8ecae6", "#219ebc")) +
  scale_y_continuous(limits = c(0, 550)) + 
  labs(
    title = "Number of LGG vs GBM Cases",
    x = "Glioma Grade",
    y = "Number of Patients"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```
LGG cases are more frequent than GBM cases in our sample,
which may influence class imbalance for downstream modeling.

```{r Age at Diagnosis by Grade}
GliomaINFO %>%
  mutate(Grade = factor(Grade, labels = c("LGG", "GBM"))) %>%
  ggplot(aes(x = Grade, y = Age_at_diagnosis, fill = Grade)) +
  geom_boxplot(width = 0.6, alpha = 0.85) +
  scale_fill_manual(values = c("#dda15e", "#bc6c25")) +
  labs(
    title = "Age Distribution by Glioma Grade",
    x = "Glioma Grade",
    y = "Age at Diagnosis (years)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```
This boxplot shows the age distribution for LGG and GBM patients.
We can see that LGG patients are generally younger, with a median age in the early 40s.
GBM patients are older on average, with a median close to 60.
This matches what we usually see in glioma cases and shows a clear age difference between the two groups before we look at genetic mutations.

```{r Mutation Frequencies Across 20 Genes}
mut_long <- GliomaINFO %>%
  select(IDH1:PDGFRA) %>%
  pivot_longer(cols = everything(),
               names_to = "Gene",
               values_to = "Mutated")

mut_freq <- mut_long %>%
  group_by(Gene) %>%
  summarize(
    count = sum(Mutated),          # number of mutated samples
    freq  = mean(Mutated),         # percentage mutated
    .groups = "drop"
  )

ggplot(mut_freq, aes(x = reorder(Gene, freq), y = freq)) +
  geom_col(fill = "#2c7fb8", width = 0.6) +
  coord_flip() +
  geom_text(
    aes(label = count, y = freq + 0.02),   # position slightly above the bar
    size = 3,
    fontface = "bold",
    color = "#1d5e8a"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Mutation Frequencies of 20 \n Glioma-Associated Genes",
    x = "Gene",
    y = "Mutation Frequency"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5))
```
This figure shows the mutation frequency of 20 glioma-associated genes.
IDH1, TP53, and ATRX are the most frequently mutated genes in our dataset, which matches what is usually seen in glioma biology.
Many other genes show much lower mutation rates, which also explains why some chi-square tests required Fisher’s exact test due to small expected counts.

```{r}
GliomaINFO %>%
  ggplot(aes(x = Age_at_diagnosis)) +
  geom_histogram(aes(y = after_stat(density), fill = factor(Grade))
                 , binwidth = 5, color = "white", alpha = 0.8) +
  geom_density(
    aes(color = factor(Grade)),  # also color by Grade
    linewidth = 1.2
  ) +
  labs(
    title = "Age Distribution with Density Curve",
    x = "Age at Diagnosis (years)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```
```{r}
shapiro.test(GliomaINFO$Age_at_diagnosis)
```
The Shapiro–Wilk test (p < 0.001) indicates that age at diagnosis is not normally distributed.
This matches the density plot, which shows a right-skewed and multimodal distribution due to younger LGG patients and older GBM patients.

## ESTABLISH COLINEARITY
```{r}
threshold <- 0.5
corr_mat <- cor(GliomaINFO, use = "pairwise.complete.obs")

high_corr_pairs <- which(abs(corr_mat) > threshold & lower.tri(corr_mat)
                         , arr.ind = TRUE)

pairs_df <- data.frame(
  Var1 = rownames(corr_mat)[high_corr_pairs[,1]],
  Var2 = colnames(corr_mat)[high_corr_pairs[,2]],
  Correlation = corr_mat[high_corr_pairs]
)

pairs_df
```

```{r warning=FALSE}
library(ComplexUpset)
library(ggplot2)

mut_cols <- c("IDH1", "PTEN", "EGFR", "NF1"
              , "PIK3R1", "FUBP1", "RB1", "SMARCA4"
              , "GRIN2A", "IDH2", "PDGFRA")  # replace with your columns
GliomaINFO$Grade <- factor(
  GliomaINFO$Grade,
  levels = c("0", "1"),
  labels = c("LGG", "GBM")
)

GliomaINFO %>%
  upset(
    mut_cols,
    base_annotations = list(
      'Intersection size' = (
        intersection_size(
          counts = FALSE,
        mapping = aes(fill = Grade)
        )
      )
      ),
  matrix = intersection_matrix(
      geom = geom_point(size = 1)
    ),
    themes = upset_modify_themes(
      theme(
        panel.spacing = unit(14, "pt"),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
        plot.margin = margin(t = -10, b = 10)
      )
    )
  )
```

More of the cases with just an IDH1 mutation have LGG, while the groups having no mutation or just PTEN or just EGFR have more GBM diagnosis. 

```{r}
tbl_summary(GliomaData %>% select(-c(Case_ID, Age_at_diagnosis, Project
                                     , Primary_Diagnosis)), by = Grade)
# Add p-value to tbl
```

Because ATRX has a stronger correlation with Grade then TP53 would, we will remove TP53 from the model. 
```{r}
muts <- GliomaData %>% 
  select(IDH1:PDGFRA) %>%
  colnames()

for (m in muts) {
  cat("\n===== Testing:", m, "=====\n")
  
  tab <- table(GliomaData$Grade, GliomaData[[m]])
  print(tab)
  
  exp_counts <- chisq.test(tab)$expected
  cat("\nExpected counts:\n")
  print(round(exp_counts, 2))
  
  # Check assumption
  if (any(exp_counts < 1) || mean(exp_counts < 5) > 0.2) {
    cat("\n Chi-square assumptions NOT met -> using Fisher's exact test.\n")
    print(fisher.test(tab))
  } else {
    cat("\n Chi-square assumptions OK.\n")
    print(chisq.test(tab))
  }
}
```
The following mutations do not show a significant difference between the two Glioma grades: PIK3CA, BCOR, CSMD3, FAT4
We can omit them from the clustering model. 

## Logistic Regression Model 
```{r}
removee <- c('PIK3CA', 'BCOR', 'CSMD3', 'FAT4', 'TP53'
             , 'Gender', 'Race', 'ATRX', 'CIC', 'MUC16', 'NOTCH1')

mData <- GliomaINFO %>%
  select(-all_of(removee)) 
```

```{r}
model <- glm(Grade ~ .,
             data = mData,
             family = binomial)
summary(model)

```


